---
title: "Hands-on Exercise 1 (Student Submission)"
author: "Cao Jinming"
date: "20 April 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format: html
---

## Learning outcome

## Getting started

### Install and load the required libraries

```{r}
pacman::p_load(tidyverse)
```

### Import the data

```{r}
exam_data <- read_csv("../data/wk1/Exam_data.csv")
```

## Introducing ggplot2

### Base R graphics versus ggplot2

```{r}
hist(exam_data$MATHS)
```

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10,
                 boundary = 100,
                 color="black",
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## Grammar of graphics

### A layered grammar of graphics

## Essential grammatical elements in ggplot2: data

```{r}
ggplot(data=exam_data)
```

## Essential grammatical elements in ggplot2: aesthetic mappings

```{r}
ggplot(data=exam_data,
       aes(x= MATHS))
```

## Essential grammatical elements in ggplot2: geoms

### Geometric objects: geom_bar

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar()
```

### Geometric objects: geom_dotplot

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,
               dotsize = 0.5) +
  scale_y_continuous(NULL,
                     breaks = NULL)
```

### Geometric objects: geom_histogram()

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram()
```

### Modify a geometric object by changing geom()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")
```

### Modify a geometric object by changing aes()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           fill = GENDER)) +
  geom_histogram(bins=20,
                 color="grey30")
```

### Geometric objects: geom_density()

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_density()
```

```{r}
ggplot(data=exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density()
```

### Geometric objects: geom_boxplot

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot()
```

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### Geometric objects: geom_violin

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_violin()
```

### Geometric objects: geom_point()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           y=ENGLISH)) +
  geom_point()
```

### Combine geom objects

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot() +
  geom_point(position="jitter",
             size = 0.5)
```

## Essential grammatical elements in ggplot2: stats

### Working with stat()

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

### Working with stat: the stat_summary() method

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour ="red",
               size=4)
```

### Working with stat: the geom() method

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",
             fun="mean",
             colour="red",
             size=4)
```

### Adding a best-fit curve to a scatterplot

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm,
              linewidth=0.5)
```

## Essential grammatical elements in ggplot2: facets

### Working with facet_wrap()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### facet_grid()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## Essential grammatical elements in ggplot2: coordinates

### Working with coordinate systems

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar()
```

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### Changing the y- and x-axis ranges

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm,
              size=0.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## Essential grammatical elements in ggplot2: themes

### Working with themes

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

## Reference

Reference: https://r4va.netlify.app/chap01
