---
title: "Hands-on Exercise 1"
author: "Cao Jinming"
date: "20 April 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format: html
---

This file follows Chapter 1 of *R for Visual Analytics*. I kept the chapter order, then explained each code line in plain English so the logic stays easy to follow.

The exam file is read from `../data/wk1/Exam_data.csv` because this exercise file lives inside the `HO_Ex` folder.

## 1.1 Learning Outcome

This chapter shows the basic building blocks of ggplot2 and how to combine them step by step.

## 1.2 Getting started

### 1.2.1 Installing and loading the required libraries

```{r}
pacman::p_load(tidyverse)
```

Layman explanation:
Line 1: `pacman::p_load(tidyverse)` — Loads the tidyverse package family in one step so R is ready for data work and plotting.

### 1.2.2 Importing data

The exam file contains student scores plus basic categories such as class, gender, and race.

```{r}
exam_data <- read_csv("../data/wk1/Exam_data.csv")
```

Layman explanation:
Line 1: `exam_data <- read_csv("../data/wk1/Exam_data.csv")` — Reads the CSV file into a table called exam_data so we can use it later.

## 1.3 Introducing ggplot

### 1.3.1 R Graphics VS ggplot

Base R gives a quick one-shot chart, while ggplot2 builds the chart in layers and gives more control.

```{r}
hist(exam_data$MATHS)
```

Layman explanation:
Line 1: `hist(exam_data$MATHS)` — Uses base R to draw a quick histogram of the Maths scores.

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10,
                 boundary = 100,
                 color="black",
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

Layman explanation:
Line 1: `ggplot(data=exam_data, aes(x = MATHS)) +` — Starts a ggplot using the exam data and maps MATHS to the x-axis. The plus sign means another layer follows.
Line 2: `geom_histogram(bins=10,` — This splits the Maths scores into 10 bins so the histogram has 10 groups.
Line 3: `boundary = 100,` — This sets the bin boundary around 100 so the histogram starts at a sensible edge.
Line 4: `color="black",` — This gives the bars a black outline.
Line 5: `fill="grey") +` — This fills the bars with grey. The plus sign means another layer follows.
Line 6: `ggtitle("Distribution of Maths scores")` — Adds a title to the plot.

## 1.4 Grammar of Graphics

Think of a plot as data plus visual mappings plus shapes, with extra layers for summaries, panels, axes, and styling.

### 1.4.1 A Layered Grammar of Graphics

ggplot2 follows the layered grammar idea: build the chart piece by piece instead of trying to draw everything in one command.

## 1.5 Essential Grammatical Elements in ggplot2: data

A ggplot object starts with the data table, even before any shape is drawn.

```{r}
ggplot(data=exam_data)
```

Layman explanation:
Line 1: `ggplot(data=exam_data)` — Starts a ggplot using the exam data as a blank plotting canvas.

## 1.6 Essential Grammatical Elements in ggplot2: Aesthetic mappings

Aesthetic mapping tells ggplot which data values should control position, colour, fill, size, and other visual features.

```{r}
ggplot(data=exam_data,
       aes(x= MATHS))
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS))` — Maps MATHS to the x-axis.

## 1.7 Essential Grammatical Elements in ggplot2: geom

Geoms are the visible marks, such as bars, dots, lines, and boxes.

### 1.7.1 Geometric Objects: geom_bar

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar()` — Counts how many students are in each category and draws one bar per category.

### 1.7.2 Geometric Objects: geom_dotplot

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x = MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_dotplot(dotsize = 0.5)` — This keeps the points small.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,
               dotsize = 0.5) +
  scale_y_continuous(NULL,
                     breaks = NULL)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x = MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_dotplot(binwidth=2.5,` — This groups the dots in 2.5-point chunks so values that are close together stay together.
Line 4: `dotsize = 0.5) +` — This keeps the points small. The plus sign means another layer follows.
Line 5: `scale_y_continuous(NULL,` — This hides the y-axis title and tick marks so the dotplot looks cleaner.
Line 6: `breaks = NULL)` — This removes the y-axis tick marks so the dotplot looks cleaner.

### 1.7.3 Geometric Objects: geom_histogram()

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x = MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_histogram()` — Draws a histogram using ggplot2’s default bin settings.

### 1.7.4 Modifying a geometric object by changing geom()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_histogram(bins=20,` — This splits the values into 20 bins so the histogram becomes more detailed.
Line 4: `color="black",` — This gives the bars a black outline.
Line 5: `fill="light blue")` — This fills the bars with light blue.

### 1.7.5 Modifying a geometric object by changing aes()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           fill = GENDER)) +
  geom_histogram(bins=20,
                 color="grey30")
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS,` — Maps MATHS to the x-axis.
Line 3: `fill = GENDER)) +` — Uses Gender to fill the histogram bars so the groups are easier to compare. The plus sign means another layer follows.
Line 4: `geom_histogram(bins=20,` — This splits the values into 20 bins so the histogram becomes more detailed.
Line 5: `color="grey30")` — This gives the bars a grey outline.

### 1.7.6 Geometric Objects: geom-density()

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_density()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x = MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_density()` — Draws a smooth density curve instead of bars to show where the values are concentrated.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x = MATHS,` — Maps MATHS to the x-axis.
Line 3: `colour = GENDER)) +` — Uses Gender to colour the curves so the groups are easy to tell apart. The plus sign means another layer follows.
Line 4: `geom_density()` — Draws a smooth density curve instead of bars to show where the values are concentrated.

### 1.7.7 Geometric Objects: geom_boxplot

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS,` — Maps MATHS to the y-axis.
Line 3: `x= GENDER)) +` — Maps GENDER to the x-axis. The plus sign means another layer follows.
Line 4: `geom_boxplot()` — Summarises each group with a boxplot that shows the median, spread, and possible outliers.

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS,` — Maps MATHS to the y-axis.
Line 3: `x= GENDER)) +` — Maps GENDER to the x-axis. The plus sign means another layer follows.
Line 4: `geom_boxplot(notch=TRUE)` — Adds notches to the boxplot so we can compare the medians more easily.

### 1.7.8 Geometric Objects: geom_violin

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_violin()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS,` — Maps MATHS to the y-axis.
Line 3: `x= GENDER)) +` — Maps GENDER to the x-axis. The plus sign means another layer follows.
Line 4: `geom_violin()` — Draws a violin plot, which shows the distribution shape in a wider-or-narrower form.

### 1.7.9 Geometric Objects: geom_point()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           y=ENGLISH)) +
  geom_point()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS,` — Maps MATHS to the x-axis.
Line 3: `y=ENGLISH)) +` — Maps ENGLISH to the y-axis. The plus sign means another layer follows.
Line 4: `geom_point()` — Draws each student as a point so we can see the relationship between two numeric variables.

### 1.7.10 geom objects can be combined

```{r}
ggplot(data=exam_data,
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot() +
  geom_point(position="jitter",
             size = 0.5)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS,` — Maps MATHS to the y-axis.
Line 3: `x= GENDER)) +` — Maps GENDER to the x-axis. The plus sign means another layer follows.
Line 4: `geom_boxplot() +` — Summarises each group with a boxplot that shows the median, spread, and possible outliers. The plus sign means another layer follows.
Line 5: `geom_point(position="jitter",` — Adds small random offsets to the points so overlapping values are easier to see, and keeps the points small.
Line 6: `size = 0.5)` — This keeps the points small.

## 1.8 Essential Grammatical Elements in ggplot2: stat

Statistics add calculated values such as means or fitted lines on top of the raw data.

### 1.8.1 Working with stat()

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS, x= GENDER)) +` — Maps MATHS to the y-axis and maps GENDER to the x-axis. The plus sign means another layer follows.
Line 3: `geom_boxplot()` — Summarises each group with a boxplot that shows the median, spread, and possible outliers.

### 1.8.2 Working with stat - the stat_summary() method

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour ="red",
               size=4)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS, x= GENDER)) +` — Maps MATHS to the y-axis and maps GENDER to the x-axis. The plus sign means another layer follows.
Line 3: `geom_boxplot() +` — Summarises each group with a boxplot that shows the median, spread, and possible outliers. The plus sign means another layer follows.
Line 4: `stat_summary(geom = "point",` — Calculates a summary value, here the mean, and draws it as a point on top of the boxplot.
Line 5: `fun = "mean",` — This tells ggplot to calculate the mean as the summary value.
Line 6: `colour ="red",` — This makes the summary point red so it stands out.
Line 7: `size=4)` — This makes the summary point larger so it is easier to notice.

### 1.8.3 Working with stat - the geom() method

```{r}
ggplot(data=exam_data,
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",
             fun="mean",
             colour="red",
             size=4)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(y = MATHS, x= GENDER)) +` — Maps MATHS to the y-axis and maps GENDER to the x-axis. The plus sign means another layer follows.
Line 3: `geom_boxplot() +` — Summarises each group with a boxplot that shows the median, spread, and possible outliers. The plus sign means another layer follows.
Line 4: `geom_point(stat="summary",` — Uses a point geom to show a summary value, here the mean, for each group.
Line 5: `fun="mean",` — This tells ggplot to calculate the mean as the summary value.
Line 6: `colour="red",` — This makes the summary point red so it stands out.
Line 7: `size=4)` — This makes the summary point larger so it is easier to notice.

### 1.8.4 Adding a best fit curve on a scatterplot?

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS, y=ENGLISH)) +` — Maps MATHS to the x-axis and maps ENGLISH to the y-axis. The plus sign means another layer follows.
Line 3: `geom_point() +` — Draws each student as a point so we can see the relationship between two numeric variables. The plus sign means another layer follows.
Line 4: `geom_smooth(size=0.5)` — This keeps the line thin.

```{r}
ggplot(data=exam_data,
       aes(x= MATHS,
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm,
              linewidth=0.5)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS,` — Maps MATHS to the x-axis.
Line 3: `y=ENGLISH)) +` — Maps ENGLISH to the y-axis. The plus sign means another layer follows.
Line 4: `geom_point() +` — Draws each student as a point so we can see the relationship between two numeric variables. The plus sign means another layer follows.
Line 5: `geom_smooth(method=lm,` — Fits a straight regression line to show the best linear trend in the data.
Line 6: `linewidth=0.5)` — This keeps the regression line thin.

## 1.9 Essential Grammatical Elements in ggplot2: Facets

Facets split one plot into several small plots so groups are easier to compare.

### 1.9.1 Working with facet_wrap()

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_histogram(bins=20) +` — This splits the values into 20 bins so the histogram becomes more detailed. The plus sign means another layer follows.
Line 4: `facet_wrap(~ CLASS)` — Splits the plot into separate small panels, one for each class.

### 1.9.2 facet_grid() function

```{r}
ggplot(data=exam_data,
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS)) +` — Maps MATHS to the x-axis. The plus sign means another layer follows.
Line 3: `geom_histogram(bins=20) +` — This splits the values into 20 bins so the histogram becomes more detailed. The plus sign means another layer follows.
Line 4: `facet_grid(~ CLASS)` — Splits the plot into a grid of panels, one for each class.

## 1.10 Essential Grammatical Elements in ggplot2: Coordinates

The chapter mentions coord_cartesian(), coord_flip(), coord_fixed(), and coord_quickmap(); here we use coord_flip() and coord_cartesian() in the examples.

### 1.10.1 Working with Coordinate

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar()` — Counts how many students are in each category and draws one bar per category.

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar() +` — Counts how many students are in each category and draws one bar per category. The plus sign means another layer follows.
Line 4: `coord_flip()` — Flips the axes so the plot becomes horizontal instead of vertical.

### 1.10.2 Changing the y- and x-axis range

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS, y=ENGLISH)) +` — Maps MATHS to the x-axis and maps ENGLISH to the y-axis. The plus sign means another layer follows.
Line 3: `geom_point() +` — Draws each student as a point so we can see the relationship between two numeric variables. The plus sign means another layer follows.
Line 4: `geom_smooth(method=lm, size=0.5)` — This keeps the line thin.

```{r}
ggplot(data=exam_data,
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm,
              size=0.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x= MATHS, y=ENGLISH)) +` — Maps MATHS to the x-axis and maps ENGLISH to the y-axis. The plus sign means another layer follows.
Line 3: `geom_point() +` — Draws each student as a point so we can see the relationship between two numeric variables. The plus sign means another layer follows.
Line 4: `geom_smooth(method=lm,` — Fits a straight regression line to show the best linear trend in the data.
Line 5: `size=0.5) +` — This keeps the line thin. The plus sign means another layer follows.
Line 6: `coord_cartesian(xlim=c(0,100),` — Zooms the view to the chosen x and y ranges without dropping the data.
Line 7: `ylim=c(0,100))` — This sets the y-axis range from 0 to 100.

## 1.11 Essential Grammatical Elements in ggplot2: themes

Themes change the look of the chart without changing the data itself.

### 1.11.1 Working with theme

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar() +` — Counts how many students are in each category and draws one bar per category. The plus sign means another layer follows.
Line 4: `coord_flip() +` — Flips the axes so the plot becomes horizontal instead of vertical. The plus sign means another layer follows.
Line 5: `theme_gray()` — Uses ggplot2’s default gray theme.

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar() +` — Counts how many students are in each category and draws one bar per category. The plus sign means another layer follows.
Line 4: `coord_flip() +` — Flips the axes so the plot becomes horizontal instead of vertical. The plus sign means another layer follows.
Line 5: `theme_classic()` — Switches to a clean classic theme with a white background.

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

Layman explanation:
Line 1: `ggplot(data=exam_data,` — Starts a ggplot using the exam data; the next line will define the visual mapping.
Line 2: `aes(x=RACE)) +` — Maps RACE to the x-axis. The plus sign means another layer follows.
Line 3: `geom_bar() +` — Counts how many students are in each category and draws one bar per category. The plus sign means another layer follows.
Line 4: `coord_flip() +` — Flips the axes so the plot becomes horizontal instead of vertical. The plus sign means another layer follows.
Line 5: `theme_minimal()` — Uses a simple minimal theme with fewer visual distractions.

## 1.12 Reference

See the online chapter for the full reference list.

In simple terms, ggplot2 builds a chart by combining data, aesthetic mappings, geoms, statistics, facets, coordinates, and themes.

Reference: https://r4va.netlify.app/chap01
