---
title: "Hands-on Exercise 7: Visual Multivariate Analysis with Heatmap"
author: "Cao Jinming"
date: "16 May 2026"
date-modified: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
format: html
---

Source chapter: [Visual Multivariate Analysis with Heatmap](https://r4va.netlify.app/chap14).

The code below follows the source chapter order. Data paths are rewritten for this repository because the exercise file lives inside `HO_Ex/`.

## Code block 1

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## Code block 2

```{r}
wh <- read_csv("../data/r4va/WHData-2018.csv")
```

## Code block 3

```{r}
row.names(wh) <- wh$Country
```

## Code block 4

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

## Code block 5

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

## Code block 6

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

## Code block 7

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

## Code block 8

```{r}
heatmaply(mtcars)
```

## Code block 9

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

## Code block 10

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

## Code block 11

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

## Code block 12

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

## Code block 13

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

## Code block 14

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

## Code block 15

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

## Code block 16

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

## Code block 17

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

## Code block 18

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

## Code block 19

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

## Code block 20

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

## Code block 21

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

## Code block 22

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```
