---
title: "Take-home Exercise 1: Visual Analytics for Data Discovery"
author: "Cao Jinming"
date: "24 May 2026"
date-modified: "last-modified"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

## Overview

This report applies visual analytics to a real-world motor insurance portfolio from a Spanish non-life insurer. The exercise brief asks for R-based visual analytics that go beyond dashboard reporting and help discover factors linked to portfolio **profitability** and **volatility**.

The report is organised in the same reproducible style as the senior reference reports: context first, analytical sandbox, data preparation, univariate exploration, bivariate analysis, statistical validation, and final findings. The analysis itself is rebuilt for the 2026 motor insurance brief and uses the Mendeley motor portfolio data.

Two analytical definitions are used throughout the report:

- **Profitability** is measured by aggregate `profit = total_premium - total_incurred` and by aggregate `loss_ratio = total_incurred / total_premium`. Lower loss ratio indicates better underwriting profitability before expenses.
- **Volatility** is measured by claim occurrence, claim frequency, capped individual loss ratio, high-loss policy rate, and upper-tail loss-ratio behaviour.

The main question is: *which portfolio segments show weaker profitability or higher volatility, and are those patterns visually and statistically defensible?*

## Analytical sandbox

### R packages and helper functions

The analytical sandbox uses `tidyverse` for data wrangling, `ggplot2` and extensions for visual analytics, and `knitr` for tabular reporting. Reusable import and portfolio-summary functions are stored in `TH_Ex/th_ex1_common.R` so that the report and slides can use the same definitions.

```{r}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)
library(knitr)
library(ggrepel)

if (file.exists("TH_Ex/th_ex1_common.R")) {
  source("TH_Ex/th_ex1_common.R")
} else {
  source("th_ex1_common.R")
}

theme_set(theme_minimal(base_size = 12))
```

### Data source and import

The dataset is the open-access motor insurance portfolio from Mendeley Data. The helper function downloads and extracts the data if the CSV and data dictionary are not already available locally. The CSV is semicolon-delimited.

```{r}
motor <- load_th1_motor()
dict <- read_excel(th1_dict_path)

glimpse(motor)
```

### Portfolio size and derived variables

The imported portfolio contains policy-year records from 2022 to 2024. Several analytical variables are derived before plotting:

- `profit`: premium minus incurred loss.
- `loss_ratio`: incurred loss divided by premium.
- `loss_ratio_cap`: individual loss ratio capped at the 99th percentile so that plots remain readable.
- `claim_flag`: whether the policy-year has at least one claim.
- `claim_frequency`: claim count divided by exposure.
- `severity`: incurred loss per claim, defined only for claim records.
- `driver_age_band` and `vehicle_age_band`: age bands for segment-level comparison.

```{r}
portfolio_size <- tibble(
  measure = c(
    "Policy-year records",
    "Distinct insured IDs",
    "Calendar years",
    "Variables after derivation",
    "Aggregate premium",
    "Aggregate incurred loss",
    "Aggregate profit",
    "Aggregate loss ratio",
    "Claim-record share"
  ),
  value = c(
    comma(nrow(motor)),
    comma(n_distinct(motor$insured_id)),
    paste(levels(motor$year), collapse = ", "),
    comma(ncol(motor)),
    money(sum(motor$total_premium, na.rm = TRUE)),
    money(sum(motor$total_incurred, na.rm = TRUE)),
    money(sum(motor$profit, na.rm = TRUE)),
    pct(sum(motor$total_incurred, na.rm = TRUE) / sum(motor$total_premium, na.rm = TRUE)),
    pct(mean(motor$claim_flag, na.rm = TRUE))
  )
)

kable(portfolio_size)
```

The portfolio is profitable in aggregate before expenses, but the aggregate result can hide segment-level volatility. The rest of the report therefore focuses on distributions, outliers, and segment comparisons rather than only aggregate totals.

### Data quality checks

```{r}
missing_tbl <- motor |>
  summarise(across(everything(), ~sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_records") |>
  filter(missing_records > 0) |>
  mutate(missing_share = missing_records / nrow(motor)) |>
  arrange(desc(missing_records))

kable(missing_tbl, digits = 4)
```

Most core premium and incurred-loss variables are complete. The largest missing count is `severity`, which is expected because severity is only defined when claims exist. Missing `fuel_type`, `vehicle_value`, `vehicle_age`, and `age_driving_licence` records are retained instead of deleted because deletion would unnecessarily reduce exposure and distort portfolio totals.

## Task 1: Univariate visual analytics

Univariate analysis is used to understand the scale, skewness, zero inflation, and category concentration of the portfolio before comparing segments.

### Numerical variables

```{r}
num_profile <- motor |>
  summarise(across(
    c(driver_age, vehicle_age, age_driving_licence, vehicle_value,
      power_to_weight_ratio, total_premium, total_claims,
      total_incurred, total_exposure, profit, loss_ratio),
    list(
      min = ~min(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      median = ~median(.x, na.rm = TRUE),
      mean = ~mean(.x, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE),
      p95 = ~quantile(.x, 0.95, na.rm = TRUE),
      max = ~max(.x, na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) |>
  pivot_longer(
    everything(),
    names_to = c("variable", ".value"),
    names_pattern = "(.+)_(min|q25|median|mean|q75|p95|max)"
  )

kable(num_profile, digits = 2)
```

```{r, fig.width=11, fig.height=7}
motor |>
  select(driver_age, vehicle_age, vehicle_value,
         total_premium, total_incurred, loss_ratio_cap) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, colour = "white", fill = "#2c7fb8") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Distributions of selected numerical variables",
    subtitle = "Individual loss ratio is capped at the 99th percentile for visual readability.",
    x = NULL,
    y = "Number of policy-year records"
  )
```

The numerical distributions are strongly skewed. Incurred loss and individual loss ratio are especially zero-inflated because most policy-year records have no claim. For this reason, medians, percentiles, and capped distributions are more informative than means alone.

### Claim count distribution

```{r, fig.width=8, fig.height=5}
motor |>
  mutate(claim_count_band = case_when(
    total_claims >= 4 ~ "4+",
    TRUE ~ as.character(total_claims)
  )) |>
  count(claim_count_band) |>
  mutate(
    claim_count_band = factor(claim_count_band, levels = c("0", "1", "2", "3", "4+")),
    share = n / sum(n)
  ) |>
  ggplot(aes(x = claim_count_band, y = share)) +
  geom_col(fill = "#41ab5d") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Most policy-year records have no claims",
    x = "Number of claims in policy year",
    y = "Share of records"
  )
```

The claim-count distribution confirms that volatility comes from a minority of claim records. This is typical for insurance portfolios: a small upper tail can strongly affect aggregate profitability.

### Categorical variables

```{r, fig.width=11, fig.height=8}
cat_vars <- c(
  "policy_type", "policy_status", "business_type", "payment_frequency",
  "bonus_score", "fuel_type", "municipality_type", "circulation_area"
)

motor |>
  select(all_of(cat_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "category") |>
  mutate(category = fct_na_value_to_level(as.factor(category), level = "Missing")) |>
  count(variable, category) |>
  group_by(variable) |>
  mutate(
    share = n / sum(n),
    category = fct_reorder(category, share)
  ) |>
  ungroup() |>
  ggplot(aes(x = category, y = share)) +
  geom_col(fill = "#756bb1") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Composition of selected categorical variables",
    x = NULL,
    y = "Share of records"
  )
```

The portfolio is not evenly distributed across categories. Some policy types, payment frequencies, and bonus-score classes dominate record count and exposure. This matters for interpretation: a small segment with a high loss ratio may be operationally important, but it must be judged together with exposure size and claim count.

## Task 2: Bivariate visual analytics and validation

Bivariate analysis compares profitability and volatility across time and major risk-related segments. The main profitability visual is aggregate loss ratio; the main volatility visuals are claim frequency, high-loss rate, and spread of capped individual loss ratios.

### Temporal trend: profitability worsens in 2024

```{r}
yearly <- portfolio_summary(motor, year)
kable(yearly, digits = 3)
```

```{r, fig.width=11, fig.height=5}
trend_amount <- yearly |>
  select(year, Premium = premium, Incurred = incurred, Profit = profit) |>
  pivot_longer(-year, names_to = "measure", values_to = "amount") |>
  ggplot(aes(x = year, y = amount, colour = measure, group = measure)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = money) +
  labs(
    title = "Premium, incurred loss, and profit by year",
    x = NULL,
    y = "Aggregate amount"
  )

trend_lr <- yearly |>
  ggplot(aes(x = year, y = loss_ratio, group = 1)) +
  geom_line(linewidth = 1, colour = "#d95f0e") +
  geom_point(size = 3, colour = "#d95f0e") +
  scale_y_continuous(labels = pct) +
  labs(
    title = "Aggregate loss ratio by year",
    x = NULL,
    y = "Loss ratio"
  )

trend_amount + trend_lr + plot_layout(widths = c(1.4, 1))
```

Premium and incurred losses both increase over time, but incurred losses grow faster in 2024. The year-level loss ratio therefore rises in 2024, indicating weaker profitability even though the overall portfolio remains profitable before expenses.

### Policy type: comprehensive covers carry higher risk

```{r}
policy_tbl <- portfolio_summary(motor, policy_type) |>
  arrange(desc(loss_ratio))

kable(policy_tbl, digits = 3)
```

```{r, fig.width=11, fig.height=6}
policy_lr <- policy_tbl |>
  mutate(policy_type = fct_reorder(policy_type, loss_ratio)) |>
  ggplot(aes(x = policy_type, y = loss_ratio, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  labs(
    title = "Aggregate loss ratio by policy type",
    subtitle = "Dashed line marks break-even before expenses.",
    x = NULL,
    y = "Loss ratio"
  )

policy_box <- motor |>
  filter(!is.na(policy_type), !is.na(loss_ratio_cap)) |>
  ggplot(aes(x = fct_reorder(policy_type, loss_ratio_cap, .fun = median), y = loss_ratio_cap)) +
  geom_boxplot(fill = "#9ecae1", outlier.alpha = 0.06) +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  labs(
    title = "Spread of capped individual loss ratio",
    subtitle = "Capping exposes group spread without letting extreme outliers dominate.",
    x = NULL,
    y = "Capped individual loss ratio"
  )

policy_lr + policy_box
```

`COMP_N` has the weakest aggregate profitability among policy types, followed by `COMP_E`. The boxplot shows that the typical policy-year still has a zero or low loss ratio, so the adverse result is driven by claim frequency and the upper tail rather than by every policy in the segment performing poorly.

### Segment scan: bonus score, fuel type, and circulation area

```{r, fig.width=11, fig.height=7}
seg_long <- bind_rows(
  portfolio_summary(motor, bonus_score) |>
    mutate(segment = "Bonus score", category = as.character(bonus_score)),
  portfolio_summary(motor, fuel_type) |>
    mutate(segment = "Fuel type", category = fct_na_value_to_level(as.factor(fuel_type), "Missing") |> as.character()),
  portfolio_summary(motor, circulation_area) |>
    mutate(segment = "Circulation area", category = as.character(circulation_area)),
  portfolio_summary(motor, payment_frequency) |>
    mutate(segment = "Payment frequency", category = as.character(payment_frequency))
)

seg_long |>
  mutate(category = fct_reorder(category, loss_ratio)) |>
  ggplot(aes(x = category, y = loss_ratio, fill = segment)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  facet_wrap(~ segment, scales = "free_y") +
  labs(
    title = "Loss ratio differs across risk-related categories",
    x = NULL,
    y = "Aggregate loss ratio"
  )
```

Non-good bonus scores, diesel vehicles, urban circulation, and non-annual payment frequencies show higher loss ratios. These signals are useful for underwriting review, but they should not be treated as causal by themselves because product mix, exposure, and vehicle profile can differ across the same segments.

### Driver and vehicle age bands

```{r, fig.width=11, fig.height=5}
age_driver <- portfolio_summary(motor, driver_age_band) |>
  mutate(segment = "Driver age", category = as.character(driver_age_band))
age_vehicle <- portfolio_summary(motor, vehicle_age_band) |>
  mutate(segment = "Vehicle age", category = as.character(vehicle_age_band))

bind_rows(age_driver, age_vehicle) |>
  filter(!is.na(category)) |>
  ggplot(aes(x = category, y = loss_ratio, group = segment, colour = segment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = pct) +
  facet_wrap(~ segment, scales = "free_x") +
  labs(
    title = "Loss ratio by driver-age and vehicle-age bands",
    x = NULL,
    y = "Aggregate loss ratio"
  ) +
  theme(legend.position = "none")
```

Age effects are visible but not perfectly monotonic. Driver age and vehicle age should therefore be interpreted as segmentation variables rather than as standalone explanations of risk.

### Policy type by year heatmap

```{r, fig.width=9, fig.height=5}
motor |>
  group_by(year, policy_type) |>
  summarise(
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    loss_ratio = incurred / premium,
    .groups = "drop"
  ) |>
  ggplot(aes(x = year, y = policy_type, fill = loss_ratio)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = pct(loss_ratio)), colour = "white", fontface = "bold") +
  scale_fill_gradient(low = "#2c7fb8", high = "#d7301f", labels = pct) +
  labs(
    title = "Loss-ratio heatmap by year and policy type",
    subtitle = "Darker red cells identify combinations needing closer monitoring.",
    x = NULL,
    y = NULL,
    fill = "Loss ratio"
  )
```

The heatmap connects the temporal and policy-type analyses. It helps identify whether a weak policy segment is persistent or concentrated in a specific calendar year.

### Brand-level volatility screen

```{r, fig.width=10, fig.height=6}
brand_tbl <- motor |>
  group_by(vehicle_brand) |>
  summarise(
    records = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    high_loss_rate = mean(loss_ratio > 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(exposure >= 1000)

brand_tbl |>
  ggplot(aes(x = claim_frequency, y = loss_ratio, size = exposure)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.65, colour = "#08519c") +
  geom_text_repel(
    data = brand_tbl |> slice_max(loss_ratio, n = 8),
    aes(label = vehicle_brand),
    size = 3,
    max.overlaps = 20
  ) +
  scale_x_continuous(labels = number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = pct) +
  scale_size_continuous(labels = comma, range = c(2, 12)) +
  labs(
    title = "Brand-level screen for high loss ratio and claim frequency",
    subtitle = "Only brands with exposure >= 1,000 are shown.",
    x = "Claim frequency",
    y = "Aggregate loss ratio",
    size = "Exposure"
  )
```

This bubble chart is a screening tool. Brands in the upper-right area combine higher claim frequency and higher loss ratio. They should be prioritised for deeper investigation, especially when the exposure bubble is large.

### Statistical validation

The dataset is large and skewed, so visual signals are validated with non-parametric and categorical tests:

- **Kruskal-Wallis tests** check whether capped individual loss ratio differs across segment groups.
- **Chi-square tests** check whether claim occurrence differs across segment groups.
- **Spearman correlations** check monotonic relationships between numerical variables and capped loss ratio.

```{r}
test_vars <- c(
  "policy_type", "bonus_score", "driver_age_band", "vehicle_age_band",
  "fuel_type", "circulation_area", "payment_frequency", "year"
)

kw_tbl <- map_dfr(test_vars, \(v) {
  kt <- kruskal.test(as.formula(paste("loss_ratio_cap ~", v)), data = motor)
  tibble(
    test = "Kruskal-Wallis",
    variable = v,
    statistic = unname(kt$statistic),
    p_value = kt$p.value
  )
})

chi_tbl <- map_dfr(test_vars, \(v) {
  ct <- chisq.test(table(motor[[v]], motor$claim_flag))
  tibble(
    test = "Chi-square",
    variable = v,
    statistic = unname(ct$statistic),
    p_value = ct$p.value
  )
})

validation_tbl <- bind_rows(kw_tbl, chi_tbl) |>
  mutate(p_value_label = format.pval(p_value, digits = 3)) |>
  select(test, variable, statistic, p_value_label)

kable(validation_tbl, digits = 2)
```

```{r}
corr_tbl <- map_dfr(
  c("driver_age", "vehicle_age", "age_driving_licence", "vehicle_value",
    "seats", "power_to_weight_ratio", "total_premium"),
  \(v) {
    ct <- suppressWarnings(cor.test(motor[[v]], motor$loss_ratio_cap,
                                    method = "spearman", exact = FALSE))
    tibble(variable = v, spearman_rho = unname(ct$estimate), p_value = ct$p.value)
  }
) |>
  mutate(p_value_label = format.pval(p_value, digits = 3)) |>
  arrange(desc(abs(spearman_rho)))

kable(corr_tbl |> select(variable, spearman_rho, p_value_label), digits = 3)
```

Most statistical tests are significant because the dataset is large. The practical interpretation should therefore come from the combination of visual difference, exposure, effect direction, and business relevance. The strongest numerical monotonic relationship is with total premium, while most demographic and vehicle correlations are small.

## Key findings

1. **The portfolio is profitable in aggregate but exposed to a claim-driven upper tail.** Most records have no claim, but a minority of claim records strongly affects incurred loss and loss ratio.
2. **Profitability weakens in 2024.** The year-level loss ratio increases, showing that incurred losses grew faster than premium.
3. **Policy type is the clearest segmentation signal.** `COMP_N` and `COMP_E` have higher loss ratios than third-party-focused products, suggesting that comprehensive cover requires closer pricing and claim review.
4. **Bonus score, fuel type, circulation area, and payment frequency show meaningful differences.** These variables help explain portfolio heterogeneity, but each should be interpreted together with exposure and product mix.
5. **Volatility is concentrated in tails, not medians.** Boxplots, capped loss-ratio charts, and the brand bubble chart show why underwriting decisions should account for high-loss policy-year behaviour.

## Recommendations

- Monitor `COMP_N` and `COMP_E` separately from the rest of the portfolio because their loss-ratio and volatility profile differs materially.
- Use the 2024 deterioration as an early-warning signal and review whether it is caused by exposure growth, claim frequency, severity, or a changing product mix.
- Treat high-loss brand, fuel, and circulation-area segments as screening candidates for deeper underwriting review rather than as immediate pricing rules.
- Extend the analysis with external enrichment, such as regional traffic density, weather exposure, or vehicle repair-cost indicators, to better explain the root causes of underperformance.

## References

- ISSS608 Take-home Exercise 1 brief: <https://isss608-ay2025-26apr.netlify.app/th_ex/th_ex1>
- Motor insurance portfolio data: <https://data.mendeley.com/datasets/sw4jmdb2sm/1>
- Senior reference examples used for report structure and visual-analytics presentation style:
  - Kylie Tan Jing Yi: <https://akalestale.netlify.app/take-home_ex/take-home_ex01/take-home_ex01>
  - Toa Zi Ying Janet: <https://janettoaisss608.netlify.app/take-home_ex/take-home_ex01/take-home_ex01>
  - Lew Ying Zhen Serena: <https://isss608-spacebun.netlify.app/take-home_ex/take-home_ex1/take-home_ex1>
  - Seng Jing Yi: <https://jingyi-vaa.netlify.app/take-home_ex/take-home_ex01/take-home_ex01>
  - Jaya George: <https://isss608-jayageorge.netlify.app/take-home_ex/take-home_ex01/take-home_ex01>
