---
title: "Take-home Exercise 1"
subtitle: "Visual Analytics for Motor Insurance Portfolio Discovery"
author: "Cao Jinming"
date: "24 May 2026"
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
---

```{r}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)
if (file.exists("TH_Ex/th_ex1_common.R")) {
  source("TH_Ex/th_ex1_common.R")
} else {
  source("th_ex1_common.R")
}
motor <- load_th1_motor()
yearly <- portfolio_summary(motor, year)
policy_tbl <- portfolio_summary(motor, policy_type) |> arrange(desc(loss_ratio))
```

## Contents

1. Analytical objective
2. Data and derived measures
3. Univariate discovery
4. Bivariate findings
5. Statistical validation
6. Implications and next steps

::: {.notes}
This presentation summarises the technical report. The focus is not only on dashboard-style reporting, but on using visual analytics to discover distributional patterns and potential drivers of profitability and volatility.
:::

## Analytical objective

- Explore numerical and categorical distributions.
- Investigate factors associated with profitability.
- Use visual evidence and statistical validation together.
- Treat extreme losses as part of volatility, not noise to ignore.

::: {.notes}
The assignment asks for both univariate and bivariate visual analytics. I define profitability through profit and loss ratio, and volatility through spread, claim occurrence, and extreme incurred-loss behaviour.
:::

## Data and sandbox

```{r}
tibble(
  measure = c("Records", "Distinct insured IDs", "Period", "Original variables"),
  value = c(nrow(motor), n_distinct(motor$insured_id), paste(levels(motor$year), collapse = "-"), 47)
) |> knitr::kable()
```

- Data: large policy-year portfolio from Mendeley motor insurance data.
- Measures: profit, loss ratio, claim flag, frequency, and severity.
- Use: discover profitability and volatility patterns across segments.

::: {.notes}
The dataset is large enough for both descriptive and statistical analysis. I use reproducible R code so the results can be regenerated from the raw Mendeley files.
:::

## Portfolio trend

```{r, fig.width=9, fig.height=4.8}
yearly |>
  select(year, premium, incurred, profit) |>
  pivot_longer(-year, names_to = "measure", values_to = "amount") |>
  ggplot(aes(year, amount, colour = measure, group = measure)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = money) +
  labs(x = NULL, y = "Aggregate amount", title = "Premium and incurred losses both increased") +
  theme_minimal()
```

- Premium and incurred losses both increased over time.
- Incurred loss growth needs to be compared against premium growth.

::: {.notes}
Premium and incurred losses both increase from 2022 to 2024. The key question is whether premium growth is sufficient to absorb the increasing loss burden.
:::

## Loss ratio worsened in 2024

```{r, fig.width=8, fig.height=4.8}
yearly |>
  ggplot(aes(year, loss_ratio, group = 1)) +
  geom_line(linewidth = 1, colour = "#d95f0e") +
  geom_point(size = 3, colour = "#d95f0e") +
  scale_y_continuous(labels = pct) +
  labs(x = NULL, y = "Aggregate loss ratio", title = "Aggregate loss ratio rose in 2024") +
  theme_minimal()
```

- 2024 has a visibly higher aggregate loss ratio.
- This suggests weaker underwriting profitability before expenses.

::: {.notes}
The 2024 loss ratio is visibly higher than 2022 and 2023. This points to declining profitability, assuming expenses are not separately modelled here.
:::

## Univariate: numerical skewness

```{r, fig.width=10, fig.height=5.6}
motor |>
  select(driver_age, vehicle_age, total_premium, total_incurred, loss_ratio_cap) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 35, colour = "white", fill = "#2c7fb8") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Records", title = "Numerical variables show strong skewness") +
  theme_minimal()
```

- Numerical variables are strongly right-skewed.
- Most records have no incurred loss, but a small upper tail is material.

::: {.notes}
The most important univariate pattern is claim zero-inflation and right skewness. Many records have no incurred loss, but a small number of records have very large loss ratios.
:::

## Univariate: categorical concentration

```{r, fig.width=10, fig.height=5.8}
motor |>
  select(policy_type, payment_frequency, bonus_score, circulation_area) |>
  pivot_longer(everything(), names_to = "variable", values_to = "category") |>
  mutate(category = fct_explicit_na(as.factor(category), "Missing")) |>
  count(variable, category) |>
  group_by(variable) |>
  mutate(pct_share = n / sum(n), category = fct_reorder(category, pct_share)) |>
  ungroup() |>
  ggplot(aes(category, pct_share)) +
  geom_col(fill = "#41ab5d") +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(x = NULL, y = "Share", title = "Portfolio mix is concentrated in selected categories") +
  theme_minimal()
```

- Portfolio mix is concentrated in selected policy and bonus-score categories.
- High loss ratios should be read together with exposure size.

::: {.notes}
Policy mix and bonus-score mix are concentrated. This affects interpretation: a smaller category can have a high loss ratio, but it must be evaluated against exposure size.
:::

## Policy type and profitability

```{r, fig.width=9, fig.height=5}
policy_tbl |>
  mutate(policy_type = fct_reorder(policy_type, loss_ratio)) |>
  ggplot(aes(policy_type, loss_ratio, fill = policy_type)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  labs(x = NULL, y = "Aggregate loss ratio", title = "Comprehensive products have higher loss ratios") +
  theme_minimal()
```

- Comprehensive products have the highest aggregate loss ratios.
- `COMP_N` is the main policy type requiring closer review.

::: {.notes}
Comprehensive products, especially COMP_N, have the highest aggregate loss ratios. Third-party-only coverage has the lowest ratio in this comparison.
:::

## Volatility by policy type

```{r, fig.width=9, fig.height=5}
motor |>
  filter(!is.na(policy_type), !is.na(loss_ratio_cap)) |>
  ggplot(aes(fct_reorder(policy_type, loss_ratio_cap, .fun = median), loss_ratio_cap)) +
  geom_boxplot(fill = "#9ecae1", outlier.alpha = 0.08) +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  labs(x = NULL, y = "Capped individual loss ratio", title = "Volatility is driven by right-tail losses") +
  theme_minimal()
```

- Capped boxplots reveal spread without extreme outliers dominating the view.
- Volatility is mainly driven by right-tail policy-year losses.

::: {.notes}
The boxplot uses capped loss ratio to make the spread visible. The large right tail confirms that volatility is a material feature of this portfolio.
:::

## Other risk-related segments

```{r, fig.width=10, fig.height=5.6}
seg_long <- bind_rows(
  portfolio_summary(motor, bonus_score) |> mutate(segment = "Bonus score", category = as.character(bonus_score)),
  portfolio_summary(motor, fuel_type) |> mutate(segment = "Fuel type", category = fct_explicit_na(as.factor(fuel_type), "Missing") |> as.character()),
  portfolio_summary(motor, circulation_area) |> mutate(segment = "Circulation", category = as.character(circulation_area))
)
seg_long |>
  mutate(category = fct_reorder(category, loss_ratio)) |>
  ggplot(aes(category, loss_ratio)) +
  geom_col(fill = "#756bb1") +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  facet_wrap(~ segment, scales = "free_y") +
  labs(x = NULL, y = "Aggregate loss ratio", title = "Loss ratio varies across selected segments") +
  theme_minimal()
```

- Bonus score, fuel type, and circulation area show visible loss-ratio differences.
- These segments are screening candidates, not final causal conclusions.

::: {.notes}
Bonus score, fuel type, and circulation area all show visible differences in aggregate loss ratio. These are useful candidates for underwriting or pricing review.
:::

## Statistical validation

```{r}
test_vars <- c("policy_type", "bonus_score", "driver_age_band", "vehicle_age_band", "fuel_type", "circulation_area", "year")
kw_tbl <- map_dfr(test_vars, \(v) {
  kt <- kruskal.test(as.formula(paste("loss_ratio_cap ~", v)), data = motor)
  tibble(variable = v, test = "Kruskal", p_value = kt$p.value)
})
chi_tbl <- map_dfr(test_vars, \(v) {
  ct <- chisq.test(table(motor[[v]], motor$claim_flag))
  tibble(variable = v, test = "Chi-square", p_value = ct$p.value)
})
bind_rows(kw_tbl, chi_tbl) |>
  mutate(p_value = format.pval(p_value, digits = 3)) |>
  pivot_wider(names_from = test, values_from = p_value) |>
  knitr::kable()
```

- Segment differences are statistically significant in the selected tests.
- Because the dataset is large, practical meaning still depends on visual effect size and exposure.

::: {.notes}
The tests show statistically significant differences across the selected segments. Because the dataset is large, I interpret these results together with visual effect size and exposure.
:::

## Brand-level screen

```{r, fig.width=9, fig.height=5.2}
brand_tbl <- motor |>
  group_by(vehicle_brand) |>
  summarise(exposure = sum(total_exposure), premium = sum(total_premium), incurred = sum(total_incurred), claims = sum(total_claims), loss_ratio = incurred / premium, claim_frequency = claims / exposure, .groups = "drop") |>
  filter(exposure >= 1000)
brand_tbl |>
  ggplot(aes(claim_frequency, loss_ratio, size = exposure)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.65, colour = "#08519c") +
  geom_text(data = brand_tbl |> slice_max(loss_ratio, n = 8), aes(label = vehicle_brand), vjust = -0.8, size = 3, check_overlap = TRUE) +
  scale_y_continuous(labels = pct) +
  labs(x = "Claim frequency", y = "Loss ratio", size = "Exposure", title = "Brand screen highlights candidates for deeper review") +
  theme_minimal()
```

- Upper-right brands combine higher claim frequency and higher loss ratio.
- Larger bubbles should be prioritised because they represent more exposure.

::: {.notes}
The brand chart is exploratory. It helps identify candidates for follow-up analysis, but it should not be treated as a final causal conclusion.
:::

## Implications

- Review premium adequacy for comprehensive products.
- Investigate 2024 deterioration in loss ratio.
- Use bonus score, fuel type, and circulation area as segmentation checks.
- Examine extreme losses separately from typical-policy behaviour.
- Validate any pricing action with exposure and mix-adjusted modelling.

::: {.notes}
The visual analytics process identifies where to investigate next. It does not replace actuarial modelling, but it provides a clear evidence trail for prioritising portfolio questions.
:::
