---
title: "Take-home Exercise 1"
subtitle: "Visual Analytics for Motor Insurance Portfolio Discovery"
author: "Cao Jinming"
date: "20 May 2026"
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    include-in-header:
      text: |
        <style>
        .reveal {
          font-size: 30px;
        }

        .reveal h1.title {
          font-size: 1.55em;
        }

        .reveal h2 {
          font-size: 1.18em;
          margin-bottom: 0.32em;
        }

        .reveal p,
        .reveal li {
          font-size: 0.78em;
          line-height: 1.16;
        }

        .reveal ul,
        .reveal ol {
          margin-top: 0.28em;
          margin-bottom: 0.28em;
        }

        .reveal table {
          font-size: 0.62em;
        }

        .reveal .cell-output-display,
        .reveal figure {
          margin: 0.12em auto 0.26em;
        }

        .reveal .cell-output-display img {
          width: 100% !important;
          max-width: 100% !important;
          max-height: 70vh !important;
          object-fit: contain;
        }
        </style>
---

```{r setup}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)
library(ggrepel)
library(knitr)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  dpi = 144
)

if (file.exists("TH_Ex/th_ex1_common.R")) {
  source("TH_Ex/th_ex1_common.R")
} else {
  source("th_ex1_common.R")
}

theme_set(theme_minimal(base_size = 11))

motor <- load_th1_motor()
yearly <- portfolio_summary(motor, year)
policy_tbl <- portfolio_summary(motor, policy_type) |> arrange(desc(loss_ratio))
overall_loss_ratio <- sum(motor$total_incurred, na.rm = TRUE) / sum(motor$total_premium, na.rm = TRUE)

seg_long <- bind_rows(
  portfolio_summary(motor, bonus_score) |>
    mutate(segment = "Bonus score", category = as.character(bonus_score)),
  portfolio_summary(motor, fuel_type) |>
    mutate(segment = "Fuel type", category = fct_na_value_to_level(as.factor(fuel_type), "Missing") |> as.character()),
  portfolio_summary(motor, circulation_area) |>
    mutate(segment = "Circulation area", category = as.character(circulation_area)),
  portfolio_summary(motor, payment_frequency) |>
    mutate(segment = "Payment frequency", category = as.character(payment_frequency))
)

brand_tbl <- motor |>
  group_by(vehicle_brand) |>
  summarise(
    records = n(),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    high_loss_rate = mean(loss_ratio > 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(exposure >= 1000)
```

## Project objective

**Main question:** Which motor insurance portfolio segments show weaker profitability or higher volatility, and are those patterns visually and statistically defensible?

-   **Profitability:** `profit = total_premium - total_incurred`; `loss_ratio = total_incurred / total_premium`.
-   **Volatility:** claim occurrence, claim frequency, capped individual loss ratio, high-loss rate, and upper-tail behaviour.
-   **Approach:** combine univariate distribution checks, bivariate segment comparisons, and statistical validation.

::: notes
This deck is aligned with the technical report objective. The goal is not only to describe the portfolio, but to identify where profitability and volatility risks are concentrated.
:::

## Data and analytical setup

```{r}
tibble(
  measure = c(
    "Policy-year records", "Distinct insured IDs", "Calendar years",
    "Variables after derivation", "Aggregate premium", "Aggregate incurred loss", "Aggregate profit",
    "Aggregate loss ratio", "Claim-record share"
  ),
  value = c(
    comma(nrow(motor)),
    comma(n_distinct(motor$insured_id)),
    paste(levels(motor$year), collapse = ", "),
    comma(ncol(motor)),
    money(sum(motor$total_premium, na.rm = TRUE)),
    money(sum(motor$total_incurred, na.rm = TRUE)),
    money(sum(motor$profit, na.rm = TRUE)),
    pct(overall_loss_ratio),
    pct(mean(motor$claim_flag, na.rm = TRUE))
  )
) |>
  kable()
```

-   The portfolio is profitable overall before expenses.
-   Segment-level analysis is still needed because aggregate totals can hide volatility.

::: notes
The dataset is a large policy-year motor portfolio from the Mendeley source. Reusable R helper functions are shared with the report.
:::

## Finding 1: profitability weakened in 2024

```{r, fig.width=11, fig.height=5}
trend_amount <- yearly |>
  select(year, Premium = premium, Incurred = incurred, Profit = profit) |>
  pivot_longer(-year, names_to = "measure", values_to = "amount") |>
  ggplot(aes(x = year, y = amount, colour = measure, group = measure)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = money) +
  labs(title = "Premium, incurred loss, and profit by year", x = NULL, y = "Aggregate amount")

trend_lr <- yearly |>
  ggplot(aes(x = year, y = loss_ratio, group = 1)) +
  geom_line(linewidth = 1, colour = "#d95f0e") +
  geom_point(size = 3, colour = "#d95f0e") +
  scale_y_continuous(labels = pct) +
  labs(title = "Aggregate loss ratio", x = NULL, y = "Loss ratio")

trend_amount + trend_lr + plot_layout(widths = c(0.58, 0.42))
```

-   Loss ratio rises from about **65.6%** in 2022 to **74.7%** in 2024.
-   This is an early-warning signal: incurred losses are growing faster than premium.

::: notes
The key message is not that the portfolio is loss-making. It remains profitable before expenses, but the margin is weakening in 2024.
:::

## Finding 2: volatility is tail-driven

```{r, fig.width=10, fig.height=5.6}
motor |>
  select(total_premium, total_incurred, total_claims, loss_ratio_cap, claim_frequency) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 35, colour = "white", fill = "#2c7fb8") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(title = "Loss and claim variables are skewed and zero-inflated", x = NULL, y = "Records")
```

-   Most policy-year records have no claims.
-   A smaller positive-claim tail drives incurred loss and volatility.
-   This justifies using capped loss ratio, claim frequency, and high-loss indicators.

::: notes
The univariate view supports the report's focus on volatility: averages alone hide the small but material claim-producing tail.
:::

## Finding 3: policy type is the strongest signal

```{r, fig.width=12, fig.height=5.8}
policy_lr <- policy_tbl |>
  mutate(
    policy_type = fct_reorder(policy_type, loss_ratio),
    label = paste0(pct(loss_ratio), " | exp ", comma(round(exposure)))
  ) |>
  ggplot(aes(x = loss_ratio, y = policy_type)) +
  geom_vline(xintercept = overall_loss_ratio, linewidth = 0.8, colour = "#08519c") +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.8, colour = "red") +
  geom_segment(aes(x = overall_loss_ratio, xend = loss_ratio, yend = policy_type), colour = "grey75") +
  geom_point(aes(size = exposure), colour = "#08519c", alpha = 0.82) +
  geom_text(aes(label = label), hjust = -0.08, size = 3.1) +
  scale_x_continuous(
    labels = pct,
    limits = c(0, max(policy_tbl$loss_ratio, overall_loss_ratio, 1, na.rm = TRUE) * 1.28)
  ) +
  scale_size_continuous(labels = comma, range = c(3, 8)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Aggregate loss ratio by policy type",
    subtitle = "Blue = portfolio loss ratio; red dashed = 100% break-even.",
    x = "Aggregate loss ratio",
    y = NULL,
    size = "Exposure"
  ) +
  theme(
    plot.margin = margin(5.5, 35, 5.5, 5.5),
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 7.5)
  )

policy_box <- motor |>
  filter(!is.na(policy_type), !is.na(loss_ratio_cap)) |>
  ggplot(aes(x = fct_reorder(policy_type, loss_ratio_cap, .fun = median), y = loss_ratio_cap)) +
  geom_boxplot(fill = "#9ecae1", outlier.alpha = 0.06) +
  coord_flip() +
  scale_y_continuous(labels = pct) +
  labs(
    title = "Spread of capped individual loss ratio",
    subtitle = "Capping exposes group spread without extreme outliers dominating.",
    x = NULL,
    y = "Capped individual loss ratio"
  ) +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 7.5)
  )

policy_lr + policy_box
```

-   `COMP_N` has the highest aggregate loss ratio, about **93.9%**.
-   `COMP_E` follows at about **73.8%**, and both comprehensive covers need separate monitoring.
-   Most policy-year records remain low-loss; weak performance is concentrated in claim frequency and upper-tail outcomes.

::: notes
The blue line is the overall portfolio loss ratio. The red dashed line is the 100% break-even line before expenses.
:::

## Finding 4: 2024 `COMP_N` is the clearest hotspot

```{r, fig.width=9.5, fig.height=5.2}
motor |>
  group_by(year, policy_type) |>
  summarise(
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    loss_ratio = incurred / premium,
    .groups = "drop"
  ) |>
  ggplot(aes(x = year, y = policy_type, fill = loss_ratio)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = pct(loss_ratio)), colour = "white", fontface = "bold") +
  scale_fill_gradient(low = "#2c7fb8", high = "#d7301f", labels = pct) +
  labs(
    title = "Loss-ratio heatmap by year and policy type",
    subtitle = "Darker red cells identify combinations needing closer monitoring.",
    x = NULL,
    y = NULL,
    fill = "Loss ratio"
  )
```

-   `COMP_N` worsens across the period and reaches about **102.4%** in 2024.
-   `COMP_E` also rises in 2024, reinforcing separate comprehensive-cover monitoring.

::: notes
The heatmap makes the recommendation more specific: review comprehensive-cover performance, especially COMP_N in 2024.
:::

## Finding 5: secondary segment signals

```{r, fig.width=11, fig.height=6}
seg_long |>
  mutate(
    category = fct_reorder(category, loss_ratio),
    label = paste0(pct(loss_ratio), " | exp ", comma(round(exposure)))
  ) |>
  ggplot(aes(x = loss_ratio, y = category)) +
  geom_vline(xintercept = overall_loss_ratio, linewidth = 0.8, colour = "#08519c") +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.8, colour = "red") +
  geom_segment(aes(x = overall_loss_ratio, xend = loss_ratio, yend = category), colour = "grey75") +
  geom_point(aes(size = exposure, colour = segment), alpha = 0.82, show.legend = FALSE) +
  geom_text(aes(label = label), hjust = -0.05, size = 2.7, show.legend = FALSE) +
  scale_x_continuous(
    labels = pct,
    limits = c(0, max(seg_long$loss_ratio, overall_loss_ratio, 1, na.rm = TRUE) * 1.30)
  ) +
  scale_size_continuous(labels = comma, range = c(2.8, 7.5)) +
  facet_wrap(~ segment, scales = "free_y") +
  coord_cartesian(clip = "off") +
  labs(
    title = "Loss ratio differs across risk-related categories",
    subtitle = "Blue line marks the overall portfolio loss ratio; red dashed line marks 100% break-even before expenses.",
    x = "Aggregate loss ratio",
    y = NULL,
    size = "Exposure"
  ) +
  theme(plot.margin = margin(5.5, 55, 5.5, 5.5))
```

-   Non-good bonus scores, diesel vehicles, urban circulation, and shorter payment frequencies show weaker experience.
-   These are screening variables, not standalone causal conclusions.

::: notes
Exposure labels are important because segment materiality depends on both loss ratio and portfolio size.
:::

## Finding 6: brand-level volatility screen

```{r, fig.width=10, fig.height=5.5}
brand_tbl |>
  ggplot(aes(x = claim_frequency, y = loss_ratio, size = exposure)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.65, colour = "#08519c") +
  geom_text_repel(
    data = brand_tbl |> slice_max(loss_ratio, n = 8),
    aes(label = vehicle_brand),
    size = 3,
    max.overlaps = 20
  ) +
  scale_x_continuous(labels = number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = pct) +
  scale_size_continuous(labels = comma, range = c(2, 12)) +
  labs(
    title = "Brand-level screen for high loss ratio and claim frequency",
    subtitle = "Only brands with exposure >= 1,000 are shown.",
    x = "Claim frequency",
    y = "Aggregate loss ratio",
    size = "Exposure"
  )
```

-   Upper-right brands combine higher claim frequency and higher loss ratio.
-   Mitsubishi and Mazda have the highest aggregate loss ratios among brands with exposure at least 1,000.
-   Large upper-right bubbles should be prioritised because they combine impact with enough exposure to matter.

::: notes
This is an exploratory screen. It supports deeper investigation rather than immediate brand-level pricing action.
:::

## Statistical validation

```{r}
test_vars <- c(
  "policy_type", "bonus_score", "driver_age_band", "vehicle_age_band",
  "fuel_type", "circulation_area", "payment_frequency", "year"
)

kw_tbl <- map_dfr(test_vars, \(v) {
  kt <- kruskal.test(as.formula(paste("loss_ratio_cap ~", v)), data = motor)
  tibble(test = "Kruskal-Wallis", variable = v, statistic = unname(kt$statistic), p_value = kt$p.value)
})

chi_tbl <- map_dfr(test_vars, \(v) {
  ct <- chisq.test(table(motor[[v]], motor$claim_flag))
  tibble(test = "Chi-square", variable = v, statistic = unname(ct$statistic), p_value = ct$p.value)
})

bind_rows(kw_tbl, chi_tbl) |>
  mutate(p_value = format.pval(p_value, digits = 3)) |>
  select(test, variable, p_value) |>
  pivot_wider(names_from = test, values_from = p_value) |>
  kable()
```

-   `policy_type` is the strongest validation signal across both tests.
-   P-values support the visuals, but large-sample significance is not causal evidence.

::: notes
The tests support group differences, but the report avoids treating p-values as sufficient evidence for pricing action.
:::

## Key findings

1.  The portfolio is profitable overall, but the margin is exposed to a small claim-producing tail.
2.  Profitability deteriorates in 2024: aggregate loss ratio rises from **65.6%** to **74.7%**.
3.  Policy type is the strongest signal; `COMP_N` reaches **93.9%** overall and **102.4%** in 2024.
4.  Bonus score, fuel type, circulation area, and payment frequency explain where weaker experience concentrates.
5.  Volatility is concentrated in claim frequency and upper-tail outcomes, not the typical record.
6.  Statistical tests support the visual signals, but the evidence remains observational and non-causal.

::: notes
This slide condenses the report's key findings and keeps them tied to the project objective: profitability and volatility discovery.
:::

## Recommendations

-   **Prioritise comprehensive-cover review:** monitor `COMP_N` and `COMP_E`, especially 2024 `COMP_N`, before broad pricing changes.
-   **Use 2024 as an early-warning trigger:** build year-by-segment loss-ratio monitoring across product and risk groups.
-   **Apply segment screening, not automatic repricing:** validate brand, diesel, urban, non-good bonus-score, and short-payment signals with controls.
-   **Strengthen volatility management:** track high-loss rate, claim frequency, severity, and capped loss ratio alongside aggregate loss ratio.
-   **Enrich portfolio data before final underwriting action:** add geography, traffic, weather, theft-risk, and repair-cost indicators.

::: notes
The recommendations are practical next steps, not automatic repricing rules. They preserve the distinction between visual discovery and confirmatory underwriting analysis.
:::

## Closing message

**Visual analytics identifies where the portfolio needs attention:**

-   2024 deterioration is the main time-based warning.
-   Comprehensive cover is the main product-based warning.
-   Tail-risk metrics are needed because portfolio averages hide claim volatility.
-   The evidence should guide investigation and enrichment before pricing action.

::: notes
The closing message links the report evidence back to the project objective and explains how visual analytics supports underwriting decision-making.
:::
